# NYC Department of Homeless Services Shelter Census (2021-2026)

Perpetua Senatus

## Introduction
Homelessness remains one of the most pressing social and administrative challenges facing New York City, impacting thousands of individuals and families daily. My interest in this topic stems from my daily commute; as someone who takes the train to come to campus, I am frequently confronted with the human reality of this crisis. I often see people sleeping on the trains and encounter families with children asking for help. These experiences have left me constantly questioning why these individuals are not provided a safe place to stay for their well-being, are the shelters simply full, or are there other barriers to entry?. To better understand these questions, I chose to study the ***NYC Department of Homeless Services (DHS) Daily Report*** because it provides a transparent, high-frequency look at the scale of the city's shelter system and the specific demographics it serves. This project aims to perform detective work through exploratory data analysis to uncover trends in shelter residency. Specifically, I am interested in studying how the total population has fluctuated over the last few years, whether there are seasonal patterns in census counts, and which subgroups are most responsible for shifts in the overall population.

## Data

### Description
The data is provided by the **New York City Department of Homeless Services(DHS)** and represents an official daily census of families and individuals residing within the city's shelter system. The dataset is **updated daily** on the NYC Open Data portal. As of the latest records used in this analysis, the dataset contains **13 columns** and over 1**800 rows,** with each row representing one Date of Census from 2021 to 2026. The key variables include the **Total Individuals in Shelter** (sum of all adults and children in the system) **and Demographic Breakdowns** (Single Adult Men/Women and Families with Children).

**Dataset Link:** [DHS Daily Report- NYC Open Data](https://data.cityofnewyork.us/Social-Services/DHS-Daily-Report/k46n-sa2m/about_data)[ ](https://data.cityofnewyork.us/Social-Services/DHS-Daily-Report/k46n-sa2m/about_data)

### Missing value analysis
```{r}
#| message: false
#| warning: false
library(tidyverse)
library(ggplot2)
df <- read_csv("https://data.cityofnewyork.us/resource/k46n-sa2m.csv?$limit=5000")
nrow(df)
#head(df)
#str(df)
#summary(df)
```

```{r}
#| warning: false
#| message: false
# convert date 
library(tidyverse)
library(lubridate)
df$date_of_census <-as.Date(df$date_of_census, format = "%m/%d/%Y")
class(df$date_of_census)
range(df$date_of_census)
head(df$date_of_census)
```

**This long list of 0 confirms that all 13 variables contain zero missing values:**
```{r}
#| message: false
#| warning: false
colSums(is.na(df)) |> sort (decreasing = TRUE)
```

**Graph 1 :** This map confirms that our data is **100% complete** across all 13 variables. It shows that the DHS has a very reliable and consistent daily reporting process:
```{r}
#| message: false
#| warning: false
library(naniar)
vis_miss(df) +
  labs(title="Global Overview of Data Completeness") +
  coord_flip() 
```

***Graph 2:*** Checking each variable one by one further confirms that every key metric has **zero missing values.** We can trust that the totals are not being skewed by one group being under-reported:
```{r}
#| message: false
#| warning: false
gg_miss_var(df) +
  labs(title ="Missing values by variable",
       y ="Number of missing values") +
       ylim(0,10)
```

**Summary of patterns:** The missing value analysis reveals that the NYC DHS Shelter dataset exhibits a complete case structure, with zero missing values identified across 1000 observations analyzed. This is likely due to the data being an official daily census where each row is an aggregated total of the entire shelter system.


## Results

### Question 1: How has the total NYC shelter population evolved between 2021 and 2026? 

#### The Impact of Economic Policy and Housing Costs
This graph tracks the effectiveness of ***'Right to Shelter'*** laws and the impact of rising NYC rent prices. A steady increase suggests that housing affordability policies may not be keeping pace with demand:
```{r}
#| warning: false
#| message: false
library(lubridate)

ggplot(df, aes(x = date_of_census , y = total_individuals_in_shelter))+
  geom_line(color="grey") +
  geom_smooth(method = "loess", color = "deeppink", se = FALSE) +
  labs(title="Total NYC Shelter Population Over Time(2021-2026)", 
       x = "Year", 
       y= "Total Individuals")+
       theme_bw()

```
**Story:** After a period of relative stability in 2021, there is a massive surge starting in late 2022 that continues into 2026, with the population nearly doubling \~ from **45,000 to over 80,000** individuals. This rapid growth suggests that the pace of people entering the system significantly exceeds the city's ability to transition them into permanent housing.

#### Does the growth trajectory suggest a system reaching its capacity? /  Ridgeline Distribution by Year
This graph tracks the distribution of the daily census by year. We can see the center of gravity of the shelter system moving toward higher numbers as the crisis evolved:
```{r}
#| message: false
#| warning: false

library(ggridges)
df |>
  mutate(Year = factor(year(date_of_census))) |>
  ggplot(aes(x = total_individuals_in_shelter, y= Year, fill = Year)) +
  geom_density_ridges(alpha = 0.7, color = "white") +
  scale_fill_manual(values = c("#FFB6C1", "#FF69B4", "#FF1493", "#C71585", "#8B008B","#4B0082")) +
  labs(title="The Evolution of the Shelter Crisis(by Year)",
       x = "Total Individuals in Shelter",
       y = "Year") +
  theme_ridges()

```

**Story:** This ridgeline plot reveals that in 2021, the system consistently stayed around 45000 people. By 2024 and 2025, the peaks have flattened and migrated significantly to the right, showing that the high population has become the new normal. Basically, the density curves shifting to the right show the system reaching new baselines.
According to this article , 2023 ....

### Question 2: How has the composition of the shelter system shifted over time(2021 - 2026)?

#### Detailed Family Composition
This graph analyses the total census population by household type so that we can understand the specific drivers of the city's housing crisis. By breaking the population into three distinct demographic groups(Single Adults, Families with Children and Adult-Only Families), we can see how the fundamental composition of the shelter system has transformed over the last five years:
```{r}
#| message: false
#| warning: false

library(tidyverse)
library(lubridate)

df_total_comp <- df |>
  
  select(date_of_census, 
         total_single_adults_in_shelter, total_individuals_in_families_with_children_in_shelter_, individuals_in_adult_families_in_shelter) |>
  
  rename("Single Adults" = total_single_adults_in_shelter,
         "Families with children" = total_individuals_in_families_with_children_in_shelter_,
         "Adult-Only Families" = individuals_in_adult_families_in_shelter) |>
  
  pivot_longer(cols = -date_of_census, names_to = "Group", values_to = "Count")


ggplot(df_total_comp, aes(x = date_of_census, y = Count, fill = Group)) +
  geom_area(alpha = 0.85) +
  scale_fill_manual(values = c(
    "Single Adults" = "#FF1493",
    "Families with children" = "#FFB6C1",
    "Adult-Only Families" ="#8B1C62")) +
    
  
  labs(title = "Evolution of NYC Shelter Residency by Household Type (2021-2026)",
       x = "Year",
       y = "Total Individuals", 
  fill = "Demographic Group") +
  theme_minimal() +
  theme (
    legend.position = "bottom",
    panel.grid.minor = element_blank() 
  )
```

**Story** This stacked area chart reveals a dramatic shift in the face of NYC's shelter residency. In 2021 and early 2022, the system was relatively stable, with **Single Adults** and **Families with Children** making up similar portions of the total. However, in late 2022, **the Families with Children** began an unprecedented expansion, quickly becoming the largest group in the system and driving the total headcount above 80,000 individuals. Meanwhile, the **Adult-Only Families** represent the smallest and most stable demographic. This visualization proves the crisis has evolved from an individual issue into a massive humanitarian challenge for families, requiring entirely different resources than the city has historically provided. 



#### The Velocity Change: Total Demographic Growth
While the daily census tracks the current population, indexing the data allows us to measure the velocity of the crisis. By setting all demographic groups to a baseline of 100 at the strat of 2021, this visualization identifies which group have experienced the most explosive rates of growth over the five-year period: 
```{r}
#| message: false
#| warning: false

library(tidyverse)
library(lubridate)

df_final_growth <- df |>
  
  arrange(date_of_census) |>
  
  mutate(
    Men_Index = (single_adult_men_in_shelter / first(single_adult_men_in_shelter)) * 100,
    Women_Index = (single_adult_women_in_shelter / first(single_adult_women_in_shelter)) * 100,
    Children_Index = (total_children_in_shelter / first( total_children_in_shelter)) * 100,  
    Parents_Index = (adults_in_families_with_children_in_shelter/first(adults_in_families_with_children_in_shelter)) * 100,
AdultFamily_Index = (individuals_in_adult_families_in_shelter/ first(individuals_in_adult_families_in_shelter)) * 100) |>
   
  filter(date_of_census == max(date_of_census))|> 
  pivot_longer(cols = ends_with("Index"), names_to = "Demographic", values_to = "Growth") |>
  mutate(Demographic = fct_recode(Demographic,
                                  "Children" = "Children_Index",
                                   "Parents" = "Parents_Index", 
                                  "Single Women" = "Women_Index", 
                                  "Single Men" ="Men_Index",
                                  "Adult Families" = "AdultFamily_Index"))

ggplot(df_final_growth, aes(x = fct_reorder(Demographic, Growth, .desc = TRUE),
                            y = Growth, fill = Demographic)) +
  
  geom_col(alpha = 0.8, show.legend = FALSE) + 
  
  
  geom_hline(yintercept = 100,linetype ="dashed", color="grey50")+

  scale_fill_manual(values = c("Parents" ="#8B1C62",
                               "Children" = "#C71585",
                               "Single Women"= "#FF1493",
                                 "Single Men"="#FF69B4",
                                "Adult Families" = "#FFB6C1"))+
  
  scale_y_continuous(expand = expansion(mult = c(0, 0.1))) +
  
  labs(title = "Total Percentage Growth by Demographic Group (2021 - 2026)", 
  x = "Demographic Group",
  y = "Indexed Growth(100 Baseline)") +
  theme_classic()

```
Story: **Parents** and **Children**  have seen their numbers skyrocket, they nearly double compared to **Single Adults** and **Adult Families** who have grown at a significantly slower pace. This ranking validates the hypothesis that the current crisis is fundamentally a family and youth crisis, where the city's primary burden has shifted toward accommodating thousands of new family households rather than the other groups.



#### The Individual Gender Divide
While family residency has driven the overall surge in the shelter system, the population of single individuals remains a significant and distinct demographic. By isolating **Single Men** and **Single Women** into a faceted yearly view, we can analyze the gender gap withing the shelter system:
```{r}
#| message: false
#| warning: false
library(tidyverse)
library(lubridate)

df_gender <- df |>
  select(date_of_census, 
         single_adult_men_in_shelter,
         single_adult_women_in_shelter)|>
  
  rename("Single Men" = single_adult_men_in_shelter,
         "Single Women" = single_adult_women_in_shelter)|>
  
  pivot_longer(cols = -date_of_census, names_to = "Gender", values_to = "Count") |>
  
  mutate(Year = year(date_of_census))

ggplot(df_gender, aes(x = date_of_census, y = Count, color = Gender)) +
  geom_line(linewidth = 1.2) +
  facet_wrap(~Year, scales ="free_x", nrow =1) +
  
  scale_color_manual(values = c("Single Men" = "#FFB6C1",
                                "Single Women" = "#FF1493")) +
    
  labs(title = "The Single Adult Gender Gap",
       x = "Timeline by Year",
       y = "Total Individuals",
       color = "Demographics") +
  theme_minimal() +
    theme(legend.position = "bottom",
    strip.background = element_rect(fill = "grey95"),
    axis.text.x = element_blank())
```
**Story:**  The faceted lines show that while **Single Men** consistently outnumber **Single Women** in the shelter system, noth groups follow nearly identical trajectories. The gap between the two lines remains relatively constant through the spikes of 2023 and into 2025. This suggests that the factors driving single individuals into the shelter system (such as low-income housing or unemployment) are affecting both genders simultaneously. 

### Question 3: Does the shelter population peak during winter months? 

####  **Seasonality** 
This graph investigates whether the shelter population is higher during the winter season. By grouping the census into seasons, we can evaluate if the population spikes significantly during winter (Dec - Feb) and compare the overall distribution across all years:
```{r}
#| message: false
#| warning: false

library(lubridate)

df_seasonal <- df |>
  mutate(Month_num = month(date_of_census),
         Season = case_when(
           Month_num %in% c(12, 1, 2) ~ "Winter",
            Month_num %in% c(3, 4, 5) ~ "Spring",
            Month_num %in% c(6, 7, 8) ~ "Summer",
            Month_num %in% c(9, 10, 11) ~ "Fall"),
         
         Season = fct_relevel(Season, "Winter", "Spring", "Summer", "Fall"))

ggplot(df_seasonal, aes(x = Season, y = total_individuals_in_shelter)) +
  
geom_boxplot(fill = "#FFB6C1" , color ="#C71585", alpha = 0.5, show.legend = FALSE, outlier.alpha = 0.2) +
  
   stat_summary(fun = median, geom = "segment", 
                aes(x = as.numeric(Season) - 0.4, 
                   xend = as.numeric(Season) + 0.4, 
                   yend = after_stat(y), color = Season), 
               size = 2.5, lineend = "round", show.legend = FALSE) +
  
  
  scale_color_manual(values = c("Winter" ="#1E90FF",
                               "Spring" = "#32CD32",
                               "Summer"="#FFD700",
                               "Fall" = "#FF8C00")) +

scale_x_discrete(labels = c ("Winter\n(Dec- Feb)", "Spring\n(Mar-May)", "Summer\n(June-Aug)", "Fall\n(Sep-Nov)")) + 
  
  labs(
    title = "Shelter Census Distribution by Season(201-2026)",
    x = "Season",
    y = "Total Individuals in Shelter") +
   theme_minimal()
```
**Story:** By using boxplots to look for seasonal patterns, we see that Winter (Dec-Feb) remains the peak occupancy period for the shelter system as expected. Interestingly, Fall (Sep-Nov) shows a higher median than Spring and Summer;  this likely reflects the pre-winter surge where dropping temperatures in October and November force people to seek indoor safety before the first frost.

#### Snow days : January 2026 and February 2026
[January 25 - 26, 2026 Blizzard ](https://www.weather.gov/okx/20260125_26#:~:text=Moderate%20to%20heavy%20snowfall%20occurred,and%20evening%20on%20January%2025.)


[February 22-23, 2026 Blizzard ](https://www.weather.gov/okx/20260222_23#:~:text=The%20National%20Weather%20Service%20in,%2C%20Southern%20New%20Haven%2C%20CT%2C)
 
I vividly remember the major blizzards earlier this year when campus was closed and my classes moved online on **January 26** and **February 23** just to keep us safe. These **snow days** provide the perfect opportunity to evaluate the real-world effectiveness of the city's **Code Blue** emergency response. By isolating the census data for the windows of January 22-29 and February 19-26, we can visualize whether the shelter population spiked as homeless were pulled off the trains and streets into safety, or if the system remained flat due to capacity limits: 
```{r}
#| message: false
#| warning: false
#| 
library(tidyverse)
library(lubridate)

  jan_blizzard <- df |>
    filter(date_of_census >="2026-01-22" & date_of_census <= "2026-01-29") |>
    mutate(Event ="Jan Blizzard")
    
  feb_blizzard <- df |>
    filter(date_of_census >= "2026-02-19" & date_of_census <= "2026-02-26")|>
      mutate(Event ="Feb Blizzard")
     
  blizzard_data <- bind_rows (jan_blizzard, feb_blizzard)
  blizzard_data$Event <- factor(blizzard_data$Event, levels = c("Jan Blizzard", "Feb Blizzard"))

ggplot(blizzard_data, aes(x = date_of_census, y= total_individuals_in_shelter, group = Event)) +
  geom_line(color ="brown", size = 1) + 
  geom_point(aes(color = date_of_census %in% as.Date(c("2026-01-25", "2026-01-26", "2026-02-22","2026-02-23"))), size = 4) +
  
  scale_color_manual( name = "Day Status",
                      values = c("FALSE" = "#FFB8C1", "TRUE" = "#FF1493"),
                      labels = c("FALSE" ="Regular Days", "TRUE" ="Blizzard Days")) +
  facet_wrap(~Event, scales = "free_x") +
  scale_y_continuous(labels = scales::comma) + 
  labs (
    title ="Code Blue Impact: Before, During, and After Peak Blizzard Days (early 2026)", 
    x = "Census Date",
    y ="Total Individuals in Shelter") +
  theme_minimal() +
  theme(legend.position = "bottom",
        strip.text = element_text(face ="bold", size = 12),
        axis.text.x = element_text(angle = 45, hjust =1))
```
**Story**: The Before, During, and After analysis of the early 2026 snow emergencies reveals the Code Blue response in action. In both January and February, the shelter census spikes exactly when blizzards hit, proving that emergency outreach successfully transitions people off the streets during life-threatening cold. The subsequent drop-offs shows the system returning to its baseline, confirming that DHS effectively scales capacity to meet immediate crisis needs.The lower February peaks likely illustrates a successful transition of individuals from January's emergency beds into more permanent housing placements as the winter progressed.


**Short summary of the most revealing findings of my analysis:** 



Take extra care to clean up your graphs, ensuring that best practices for presentation are followed, as described in the audience-ready style section below.

Your project should contain 5-10 graphs. Quality is rewarded over quantity.


## Conclusion


Discuss: 
main takeaways of your exploration,  --> 
limitations, --> numbers only , high frequency daily time series,  dataset aggregates adults withing family units meaning parents category include all mothers, fathers, and guardians in the system and I wanted to have a viz about single mothers and single fathers, children are attached to family units so I cannot really isolated them 
future directions,  --> 
lessons learned. --> code blue, spike in 2023, shelter capacity

(suggested length: one paragraph)
